---
title: "AlessandroKerr_A2"
author: "Alessandro Kerr"
date: " February 1, 2022"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(tigris)
library(censusapi)
library(sf)
library(leaflet)
library(htmltools)
library(tidycensus)
library(mapview)
library(dplyr)
library(ggrepel)
library(ggplot2)

knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning = F, message = F)

```

```{r}
path <- "/Volumes/GoogleDrive/Shared drives/SFBI/Data Library/NHTS/nhts17-caltrans-tsdc-download/"

pois <- st_read("/Volumes/GoogleDrive/Shared drives/SFBI/Data Library/OSM/gis_osm_pois_a_free_1.shp")

```

```{r}
# pois_summary <- pois %>% 
#   st_drop_geometry() %>% 
#   group_by(fclass) %>% 
#   count() %>% 
#   arrange(desc(n))
# 
# bay_county_names <-
#   c(
#     "Alameda",
#     "Contra Costa",
#     "Marin",
#     "Napa",
#     "San Francisco",
#     "San Mateo",
#     "Santa Clara",
#     "Solano",
#     "Sonoma"
#   )
# 
# bay_counties <-
#   counties("CA", cb = T, progress_bar = F) %>%
#   filter(NAME %in% bay_county_names) %>%
#   st_transform(st_crs(pois))
# 
# bay_pois <- pois %>% 
#   .[bay_counties, ] %>% 
#   rename(amenity = fclass)
# 
# mapview(bay_pois, zcol = "amenity")
# 
# #Let's not use all POIs lets filter down to differentiate
# 
# bay_pois_filter <- bay_pois %>% 
#   filter(amenity %in% c(
#     "park",
#     "supermarket",
#     "pharmacy",
#     "bank",
#     "school",
#     "doctors",
#     "restaurant",
#     "bar",
#     "pub",
#     "pitch",
#     "playground",
#     "hospital",
#     "theatre"
#   ))
# saveRDS(bay_pois_filter, "bay_pois_filter.RDS")
```

```{r}
# bay_pois_filter2 <- bay_pois %>% 
#   filter(amenity %in% c(
#     "hospital"))
# 
# saveRDS(bay_pois_filter2, "bay_pois_filter2.RDS")
```

```{r}
# smc_cbs <- blocks("CA","San Mateo")
# 
# rdw_boundary <- places("CA") %>% 
#   filter(NAME == "Redwood City")
# 
# rdw_cbs <- smc_cbs %>% 
#   st_centroid() %>% 
#   .[rdw_boundary, ] %>% 
#   st_drop_geometry() %>% 
#   left_join(smc_cbs %>% select(GEOID10)) %>% 
#   st_as_sf()
# 
# saveRDS(rdw_cbs, "rdw_cbs.RDS")
```

```{r}
bay_pois_filter <- readRDS("bay_pois_filter.RDS")
bay_pois_filter2 <- readRDS("bay_pois_filter2.RDS")

mapview(bay_pois_filter, zcol = "amenity")
mapview(bay_pois_filter2, zcol = "amenity")
```

```{r}
rdw_cbs <- readRDS("rdw_cbs.RDS")
mapview(rdw_cbs)
```


```{r}
devtools::install_github("walkerke/mapboxapi")

library(mapboxapi)

mb_access_token("sk.eyJ1IjoiYWxleG5nbGQiLCJhIjoiY2t5dnY3OGliMDJlaTMwcnIxeHd5bTJucyJ9.mZzy6K59_f5dGUlEQ8oL7Q", install = T, overwrite = T)
```

```{r}
# isochrone <- mb_isochrone(
#   rdw_cbs,
#   profile = "walking",
#   time = c(5,10,15)
# ) %>% mutate(
# mode = "driving")
# 
# isochrone[1:3,] %>% mapview()
# 
#saveRDS(isochrone,"isochrone_rwc_walking.RDS")

# isochrone_2 <- mb_isochrone(
#   rdw_cbs,
#   profile = "driving",
#   time = c(5,10,15)
# ) %>% mutate(
# mode = "driving")
# 
# isochrone_2[1:3,] %>% mapview()
# 
# saveRDS(isochrone_2, "isochrone_rwc_driving.RDS") 

# isochrone_3 <- mb_isochrone(
#   rdw_cbs,
#   profile = "cycling",
#   time = c(5,10,15)
# ) %>% mutate(
#   mode = "biking")

# isochrone_3[1:3,] %>% mapview()
# 
# saveRDS(isochrone_3, "isochrone_rwc_biking.RDS")


```


Did separately on purpose so that we could share the tokens across multiple users. Otherwise would have used map_dfr function to streamline. 

```{r}

isochrone <- readRDS("isochrone_rwc_walking.RDS")
isochrone_2 <- readRDS("isochrone_rwc_driving.RDS")
isochrone_3 <- readRDS("isochrone_rwc_biking.RDS")

isochrone_combined <- rbind(isochrone, isochrone_2, isochrone_3)

# Use driving for hospital analysis. Use walking for other POIs. 
```

```{r}
access_raw <- isochrone_combined %>% 
  st_make_valid() %>%
  st_join(bay_pois_filter) %>% 
  st_drop_geometry()

```

```{r}
amenity_preference <- data.frame(
  amenity = c(
    "park",
    "supermarket",
    "pharmacy",
    "bank",
    "school",
    "doctors",
    "restaurant",
    "bar",
    "pub",
    "pitch",
    "playground",
    "hospital",
    "theatre"
  ),
  amenity_value = c(
    0.9,
    1,
    0.5,
    0.5,
    0.8,
    0.6,
    0.7,
    0.7,
    0.7,
    0.8,
    0.8,
    0.9,
    0.3
  ),
  amenity_quantity = c(
   2,
   2,
   1,
   1,
   1,
   2,
   20,
   5,
   5,
   10,
   5,
   1,
   2
  )
) %>% 
  mutate(
    amenity_decay = -log(0.5)/(amenity_quantity)
  )

mode_preference <- data.frame(
  mode = c(
    "walking",
    "biking",
    "driving"
  ),
  mode_value = c(
    1,
    0.5,
    0.7
  ),
  mode_reasonable = c(
    15,
    15,
    25
  )
) %>% 
  mutate(
    mode_decay = -log(0.5)/mode_reasonable
  )
```

```{r}
complete_baseline <- data.frame(
  amenity = amenity_preference$amenity %>% 
    rep(amenity_preference$amenity_quantity)
) %>% 
  left_join(
    amenity_preference,
    by = "amenity"
  ) %>% 
  group_by(amenity) %>% 
  mutate(
    amenity_rank = row_number() - 1
  ) %>% 
  ungroup() %>% 
  mutate(
    score = amenity_value * exp(-amenity_rank * amenity_decay) * 0.5
  ) # ASSUMES ALL 0.5, or mode_factor for 15 min of any transport means

sum(complete_baseline$score)
```

```{r}
complete_temp <- access_raw %>%
  left_join(
    amenity_preference,
    by = "amenity"
  ) %>%
  select(-amenity_quantity) %>%
  left_join(
    mode_preference,
    by = "mode"
  ) %>%
  select(-mode_reasonable) %>% 
  group_by(id,mode,amenity) %>%
  arrange(time) %>%
  mutate(
    amenity_rank = row_number() - 1
  ) %>% 
  ungroup()
```

```{r}
 complete_modes <- complete_temp %>% #for one mode only
    mutate(
      score = amenity_value *exp(-amenity_rank * amenity_decay) * exp(-time * mode_decay)
    ) %>%
    group_by(id, mode) %>%
    arrange(desc(score)) %>%
    filter(!duplicated(osm_id)) %>%
    summarize(
      score = sum(score, na.rm=T)/sum(complete_baseline$score)
    )
```

```{r}
complete_total <- complete_temp %>%
  mutate(
      score = amenity_value *exp(-amenity_rank * amenity_decay) * mode_value * exp(-time * mode_decay)
    ) %>%
  group_by(id) %>%
  arrange(desc(score)) %>%
  filter(!duplicated(osm_id)) %>%
  summarize(
    score = sum(score, na.rm = T)/sum(complete_baseline$score)
  ) %>%
  mutate(mode = "total")

complete <- rbind(
  complete_modes,
  complete_total
)

complete_map <- complete %>%
  pivot_wider(
    names_from = "mode",
    values_from = "score"
  ) %>%
  cbind(rdw_cbs %>% select(GEOID10)) %>%
  st_as_sf()

mapview(complete_map, zcol = "total")
```

# EQUITY ANALYSIS
```{r}

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

dec_vars_2020 <-
  listCensusMetadata(
    name = "2020/dec/pl",
    type = "variables"
  )

rwc_pop_race_2020 <-
  getCensus(
    name = "dec/pl",
    vintage = 2020,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "group(P1)"
  ) %>% 
  mutate(
    block =
      paste0(state,county,tract,block)
  ) %>% 
  select(!c(GEO_ID,state,county,tract,NAME) & !ends_with(c("NA"))) %>% 
  pivot_longer(
    ends_with("N"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    dec_vars_2020 %>% 
      select(name, label)
  ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,NA,"category1","category2"),
    sep = "!!"
  ) %>%
  left_join(
    rdw_cbs %>%
      select("GEOID10"),
    by = c("block"="GEOID10") 
  ) %>%
  select(-category1)

score_median = median(complete_map$total) 
  
rwc_filter <- rwc_pop_race_2020 %>%
  filter(!is.na(category2)) %>%
  left_join(
    complete_map %>%
      select("total"),
    by = 'geometry'
  ) %>%
  filter(!is.na(total)) %>%
  mutate(
    Threshold = ifelse(total>score_median,"Above Median","Below Median")
  ) %>%
  mutate(
    race = case_when(
      category2 %in% c("White alone")~"White",
      category2 %in% c("Black or African American alone")~"Black or African American",
      category2 %in% c("American Indian and Alaska Native alone")~"American Indian and Alaska Native",
      category2 %in% c("Asian alone")~"Asian",
      category2 %in% c("Native Hawaiian and Other Pacific Islander alone")~"Native Hawaiian and Other Pacific Islander",
      category2 %in% c("Some Other Race alone", "Population of two races:", "Population of three races:" , "Population of four races:" , "Population of five races:",  "Population of six races:")~"Other"
    )
  ) %>%
  group_by(
    Threshold, race
  ) %>%
  summarize(
   estimate=sum(estimate)
  ) %>%
  group_by(
    Threshold
  ) %>%
  mutate(
    perc = estimate/sum(estimate) *100
  )

df2 <- rwc_filter %>% 
  mutate(csum = rev(cumsum(rev(perc))), 
         pos = perc/2 + lead(csum, 1),
         pos = if_else(is.na(pos), perc/2, pos))

rwc_filter %>% 
  ggplot(data = rwc_filter,
         mapping = aes(x = "" , 
             y = perc, 
             fill = fct_inorder(race)
         )) +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~Threshold) +
  scale_fill_brewer(palette = "Pastel1") +
  geom_label_repel(data = df2,
                  mapping = aes(y = pos, label = paste0(round(perc), "%")),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Completeness Score by Race")) +
  theme_void()

```
# CRITICAL AMENITY ANALYSIS

```{r}
access_raw2 <- isochrone_2 %>% 
  st_make_valid() %>%
  st_join(bay_pois_filter2) %>% 
  st_drop_geometry() %>%
  filter(!is.na(osm_id)) %>%
  group_by(
    id
  ) %>%
  mutate(
    mintime = min(time)
  ) %>%
  left_join(
    complete_map%>%
      select(id,geometry),
    by="id"
  ) %>%
  filter(!duplicated(id)) %>%
  mutate(
    Access = ifelse(mintime>10,"No","Yes")
  ) %>%
  st_as_sf()

mapview(access_raw2, zcol="Access")
```

```{r}
new_hospital <- data.frame(
  osm_id=00000001, code = 2110, amenity = "hospital", name = "New Hospital")

newhospital_point <- as.data.frame(37.48956) %>%
  rename(lat = "37.48956") %>%
  mutate(lon = -122.2239) %>%
  sf::st_as_sf(., coords = c("lon", "lat"), crs = 4326)

new_hospital <- new_hospital %>%
  cbind(newhospital_point) %>%
  st_as_sf(crs=4326)
  
mapview(new_hospital)

bay_pois_filternew <- bay_pois_filter2 %>%
  rbind(new_hospital)

mapview(bay_pois_filternew)

```

```{r}

access_raw3 <- isochrone_2 %>% 
  st_make_valid() %>%
  st_join(bay_pois_filternew) %>% 
  st_drop_geometry() %>%
  filter(!is.na(osm_id)) %>%
  group_by(
    id
  ) %>%
  mutate(
    mintime = min(time)
  ) %>%
  left_join(
    complete_map%>%
      select(id,geometry),
    by="id"
  ) %>%
  filter(!duplicated(id)) %>%
  mutate(
    Access = ifelse(mintime>10,"No","Yes")
  ) %>%
  st_as_sf() 

mapview(access_raw3, zcol="Access")

```

# WALK SCORE COMPARISON

```{r}
library(walkscoreAPI)


rwc_centroid <- rdw_cbs %>%
  st_centroid() %>%
  extract(geometry, c('lat', 'lon'), '\\((.*), (.*)\\)', convert = TRUE)


getWS("longitude", "latitude", key)

```

NEXT STEPS:
- fill in code above to match our data. --> need to expand POIs and get biking isochronesx
- equity analysis (using DECENNIAL DATA)
- hospital analysis -- redo but only for hospitals and driving 
  - then add in one extra hospital and see what effect it has on communities
  
  changes:
  (1) using blocks instead of block groups (and consequently decennial data)
  (2) pick larger amount of POIs
  (3) Hospital "critical amenities" analysis - either has minimum access or not (binary)
  (4) Equity analysis
  (5) By manually adding development - see how affects completeness for small area
      - isolate area of of blocks and view before/after change. 
  (6) Compare our isochrone_1 to walkscore.api
  
  
  * Make sure to incorporate saveRDS for marks
  