---
title: "CEE218 Assignment 1"
author: "Alexander Ng"
date: "9/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```


```{r}
library(tidyverse)

year <- 2017
quarters <- 1:4
type <- "Electric"

pge_17_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_17_elec <- rbind(pge_17_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_17_elec, "pge_17_elec.rds")
}

year <- 2018
quarters <- 1:4
type <- "Electric"

pge_18_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_18_elec <- rbind(pge_18_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_18_elec, "pge_18_elec.rds")
}

year <- 2019
quarters <- 1:4
type <- "Electric"

pge_19_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_19_elec <- rbind(pge_19_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_19_elec, "pge_19_elec.rds")
}

year <- 2020
quarters <- 1:4
type <- "Electric"

pge_20_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_20_elec <- rbind(pge_20_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_20_elec, "pge_20_elec.rds")
}

year <- 2021
quarters <- 1:2
type <- "Electric"

pge_21_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_21_elec <- rbind(pge_21_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_21_elec, "pge_21_elec.rds")
}
pge_electric<- do.call("rbind", list(pge_17_elec, pge_18_elec, pge_19_elec, pge_20_elec, pge_21_elec))

saveRDS(pge_electric, "pge_electric.rds")

```

```{r}
year <- 2017
quarters <- 1:4
type <- "Gas"

pge_17_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_17_gas <- rbind(pge_17_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_17_gas, "pge_17_gas.rds")
}

year <- 2018
quarters <- 1:4
type <- "Gas"

pge_18_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_18_gas <- rbind(pge_18_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_18_gas, "pge_18_gas.rds")
}

year <- 2019
quarters <- 1:4
type <- "Gas"

pge_19_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_19_gas <- rbind(pge_19_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_19_gas, "pge_19_gas.rds")
}

year <- 2020
quarters <- 1:4
type <- "Gas"

pge_20_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_20_gas <- rbind(pge_20_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_20_gas, "pge_20_gas.rds")
}

year <- 2021
quarters <- 1:2
type <- "Gas"

pge_21_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_21_gas <- rbind(pge_21_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_21_gas, "pge_21_gas.rds")
}
pge_gas<- do.call("rbind", list(pge_17_gas, pge_18_gas, pge_19_gas, pge_20_gas, pge_21_gas))

saveRDS(pge_gas, "pge_gas.rds")

```
```{r merge}
pge_combined<-merge(pge_electric, pge_gas, all=TRUE)
pge_filter_residential<-filter(pge_combined, CUSTOMERCLASS %in% c("Elec- Residential","Gas- Residential"))
pge_filter_commercial<-filter(pge_combined, CUSTOMERCLASS %in% c("Elec- Commercial", "Gas- Commercial"))

```

```{r mutate residential}

pge_residentialfinal <-
  pge_filter_residential %>% 
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>% 
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>% 
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, 
        na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM,
        na.rm = T
      )
  ) %>% 
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41KBTU
  ) %>%
  mutate(
    GAS_KBTU = TOTALTHM*100 #1THM=100KBTU
  ) %>%
  mutate(
    DATE=
      paste(YEAR,MONTH, "01", sep="-") %>% as.Date()
  ) %>%
  mutate(TOTALBTU = ELEC_KBTU+GAS_KBTU)
```

```{r mutate commmercial}
pge_commercialfinal <-
  pge_filter_commercial %>% 
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>% 
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>% 
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, 
        na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM,
        na.rm = T
      )
  ) %>% 
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41KBTU
  ) %>%
  mutate(
    GAS_KBTU = TOTALTHM*100 #1THM=100KBTU
  ) %>%
  mutate(
    DATE=
      paste(YEAR,MONTH, "01", sep="-") %>% as.Date()
  ) %>%
  mutate(TOTALBTU = ELEC_KBTU+GAS_KBTU)
```
```{r}
library(plotly)

pge_plotresidential<-
  pge_residentialfinal %>%
  ggplot() +
  geom_bar(
    aes(
      x = DATE ,
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "dodge"
  ) +
  labs(
    x="Time", y="kBTU", 
    title = "PGE Residential Electric & Gas Consumption Data from 2017 to 2021",
    fill = "Fuel Type"
  )

pge_plotresidential %>% 
  ggplotly() %>% 
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)

```
```{r}
library(plotly)

pge_plotcommercial<-
  pge_commercialfinal %>%
  ggplot() +
  geom_bar(
    aes(
      x = DATE ,
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "dodge"
  ) +
  labs(
    x="Time", y="kBTU", 
    title = "PGE Commercial Electric & Gas Consumption Data from 2017 to 2021",
    fill = "Fuel Type"
  )

pge_plotcommercial %>% 
  ggplotly() %>% 
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)

```

## General Observations

Based on the residential and commercial energy consumption data from PGE, it would appear that the pandemic led to a slight increase in electrical consumption for residential properties and a slight decrease in electrical consumption for commercial properties (as compared to preceding years). 

This can be attributed to public health requirements for the general workforce to work from home. Based on preliminary 2021 data, this trend appears to have reversed and electrical consumption appears to be similar to the years before the pandemic, which can likely be attributed to relaxation of public health policies and a return to office arrangement for the general workforce. Interestingly, the electrical consumption at commercial properties in 2021 appears to be increasing further beyond historical averages. This could be a sign that building/facility owners have invested in upgrades to HVAC equipment (which requires more electrical consumption) to improve building ventilation as a mitigation against the spread of COVID-19.

The pandemic did not appear to have a direct impact on commercial and residential gas consumption which appears to be tied to seasonal changes in weather instead.

