---
title: "CEE218 Assignment 1_Resubmission"
author: "Alexander Ng"
date: "9/27/2021 - Revised on 10/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = F, warning = F, echo = F)
```


```{r, include=FALSE}
knitr::opts_chunk$set(message = F, warning = F, echo = F)

library(tidyverse)

year <- 2017
quarters <- 1:4
type <- "Electric"

pge_17_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  
  print(filename)
  temp <- read_csv(filename)
  pge_17_elec <- rbind(pge_17_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_17_elec, "pge_17_elec.rds")
}

pge_17_elec<-pge_17_elec[!duplicated(pge_17_elec[c('ZIPCODE','YEAR','MONTH','CUSTOMERCLASS')]),]

year <- 2018
quarters <- 1:4
type <- "Electric"

pge_18_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_18_elec <- rbind(pge_18_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_18_elec, "pge_18_elec.rds")
}

year <- 2019
quarters <- 1:4
type <- "Electric"

pge_19_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_19_elec <- rbind(pge_19_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_19_elec, "pge_19_elec.rds")
}

year <- 2020
quarters <- 1:4
type <- "Electric"

pge_20_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_20_elec <- rbind(pge_20_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_20_elec, "pge_20_elec.rds")
}

year <- 2021
quarters <- 1:2
type <- "Electric"

pge_21_elec <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Electric Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  knitr::opts_chunk$set(message = F, warning = F, echo = F)
  
  print(filename)
  temp <- read_csv(filename)
  pge_21_elec <- rbind(pge_21_elec,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_21_elec, "pge_21_elec.rds")
}
pge_electric<- do.call("rbind", list(pge_17_elec, pge_18_elec, pge_19_elec, pge_20_elec, pge_21_elec))

saveRDS(pge_electric, "pge_electric.rds")
knitr::opts_chunk$set(message = F, warning = F, echo = F)
```


```{r, include=FALSE}

knitr::opts_chunk$set(message = F, warning = F, echo = F)

year <- 2017
quarters <- 1:4
type <- "Gas"

pge_17_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_17_gas <- rbind(pge_17_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_17_gas, "pge_17_gas.rds")
}

pge_17_gas<-pge_17_gas[!duplicated(pge_17_gas[c('ZIPCODE','YEAR','MONTH','CUSTOMERCLASS')]),]

year <- 2018
quarters <- 1:4
type <- "Gas"

pge_18_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_18_gas <- rbind(pge_18_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_18_gas, "pge_18_gas.rds")
}

year <- 2019
quarters <- 1:4
type <- "Gas"

pge_19_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_19_gas <- rbind(pge_19_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_19_gas, "pge_19_gas.rds")
}

year <- 2020
quarters <- 1:4
type <- "Gas"

pge_20_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  temp <- read_csv(filename)
  pge_20_gas <- rbind(pge_20_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_20_gas, "pge_20_gas.rds")
}

year <- 2021
quarters <- 1:2
type <- "Gas"

pge_21_gas <- NULL

for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/Gas Usage/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  
  print(filename)
  temp <- read_csv(filename)
  pge_21_gas <- rbind(pge_21_gas,temp)
  # Note rbind requires field names to be consistent for every new thing that you add.
  saveRDS(pge_21_gas, "pge_21_gas.rds")
}
pge_gas<- do.call("rbind", list(pge_17_gas, pge_18_gas, pge_19_gas, pge_20_gas, pge_21_gas))

saveRDS(pge_gas, "pge_gas.rds")
knitr::opts_chunk$set(message = F, warning = F, echo = F)
```

```{r merge}
knitr::opts_chunk$set(message = F, warning = F, echo = F)

pge_combined<-merge(pge_electric, pge_gas, all=TRUE)
pge_filter_residential<-filter(pge_combined, CUSTOMERCLASS %in% c("Elec- Residential","Gas- Residential"))
pge_filter_commercial<-filter(pge_combined, CUSTOMERCLASS %in% c("Elec- Commercial", "Gas- Commercial"))

```

```{r mutate residential}
knitr::opts_chunk$set(message = F, warning = F, echo = F)

pge_residentialfinal <-
  pge_filter_residential %>% 
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>% 
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>% 
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, 
        na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM,
        na.rm = T
      )
  ) %>% 
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41KBTU
  ) %>%
  mutate(
    GAS_KBTU = TOTALTHM*100 #1THM=100KBTU
  ) %>%
  mutate(
    DATE=
      paste(YEAR,MONTH, "01", sep="-") %>% as.Date()
  ) %>%
  mutate(TOTALBTU = ELEC_KBTU+GAS_KBTU)
```

```{r mutate commmercial}
knitr::opts_chunk$set(message = F, warning = F, echo = F)

pge_commercialfinal <-
  pge_filter_commercial %>% 
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>% 
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>% 
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, 
        na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM,
        na.rm = T
      )
  ) %>% 
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41KBTU
  ) %>%
  mutate(
    GAS_KBTU = TOTALTHM*100 #1THM=100KBTU
  ) %>%
  mutate(
    DATE=
      paste(YEAR,MONTH, "01", sep="-") %>% as.Date()
  ) %>%
  mutate(TOTALBTU = ELEC_KBTU+GAS_KBTU)
```
```{r}
knitr::opts_chunk$set(message = F, warning = F, echo = F)

library(plotly)

pge_plotresidential<-
  pge_residentialfinal %>%
  ggplot() +
  geom_line(data=pge_residentialfinal, aes(x = DATE, y = TOTALBTU, colour = CUSTOMERCLASS, group = CUSTOMERCLASS)) +
  labs(
    x="Time", y="kBTU", 
    title = "Residential Electricity & Gas Consumer Data from 2017 to 2021",
    fill = "Legend"
  )

pge_plotresidential %>% 
  ggplotly() %>% 
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)

```
```{r}
knitr::opts_chunk$set(message = F, warning = F, echo = F)

library(plotly)

#ggplot(dat, aes(time_new, value)) + 
  #geom_line() + 
 # scale_x_continuous("Time", breaks=seq(0,24,3), labels=sprintf("%02d:00", seq(0,24,3))

#ggplot(data = tsla_stock_metrics, aes(x = date, y = close_price)) +
  #geom_line()

## plot
##ggplot(data = df, aes(x=x, y=val)) + geom_line(aes(colour=variable))

pge_plotcommercial<-
  pge_commercialfinal %>%
  ggplot() +
  geom_line(data=pge_commercialfinal, aes(x = DATE, y = TOTALBTU, colour = CUSTOMERCLASS, group = CUSTOMERCLASS)) +
  labs(
    x="Time", y="kBTU", 
    title = "Commercial Electricity & Gas Consumer Data from 2017 to 2021",
    fill = "Legend"
  )

pge_plotcommercial %>% 
  ggplotly() %>% 
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)

```

## General Observations

Based on the residential and commercial energy consumption data from PGE, it would appear that the pandemic led to a slight increase in electrical consumption for residential properties and a slight decrease in electrical consumption for commercial properties (as compared to preceding years). 

This can be attributed to public health requirements for the general workforce to work from home. Based on preliminary 2021 data, this trend appears to have reversed and electrical consumption appears to be similar to the years before the pandemic, which can likely be attributed to relaxation of public health policies and a return to office arrangement for the general workforce. Interestingly, the electrical consumption at commercial properties in 2021 appears to be increasing further beyond historical averages. This could be a sign that building/facility owners have invested in upgrades to HVAC equipment (which requires more electrical consumption) to improve building ventilation as a mitigation against the spread of COVID-19.

The pandemic did not appear to have a direct impact on commercial and residential gas consumption which appears to be tied to seasonal changes in weather instead.

Assumptions: 
Data supplied by PGE are reflective of energy consumption amongst the general population. (For example, there could be certain segments of the population who are off the grid or supplement their energy consumption through other sources such as renewables.)

There may also be unverified errors in the data supplied by PGE. For instance, the data registered for residential and commercial electricity consumption in Sep 2017 were significantly higher (over 50%) than other data. Therefore, further consultation with PGE is required to identify the source and cause of this outlier.


