---
title: "AlexanderNg_A8"
author: "Alexander Ng"
date: "11/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r extract bay pumas, , message=FALSE, warning=FALSE, echo=FALSE}
library(tigris)
library(tidyverse)
library(tidycensus)
library(sf)
library(censusapi)
library(leaflet)
library(StatMatch)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

```


```{r mark Antioch BART , message=FALSE, warning=FALSE, echo=FALSE}

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_pumas,
    weight = 1,
    color = "gray",
    label = ~PUMACE10
  ) %>% 
  addMarkers(
    lng = -121.7822953,
    lat = 37.9957278
  ) %>% 
  addPolygons(
    data = bay_pumas %>% 
      filter(PUMACE10 == "01309")
  )

```


```{r filter 2014-2019 data, message=FALSE, warning=FALSE, echo=FALSE}

pums_2014_2019<-readRDS('pums_2014_2019_wts.rds')

  
pums_bart <- pums_2014_2019 %>%
  mutate(
    PWGTP = as.numeric(PWGTP),
    bart = ifelse(
      JWTR %in% c("4"),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA, year) %>% 
  summarize(
    pop = sum(PWGTP),
    bart = sum(bart)
  )


```


```{r clean BART PUMAS, message=FALSE, warning=FALSE, echo=FALSE}

pums_bart_clean <-
  pums_bart %>% 
  select(-pop) %>% 
  pivot_wider(
    names_from = year,
    values_from = bart
  )

```

```{r match years, message=FALSE, warning=FALSE, echo=FALSE}
obs_matrix <-
  pums_bart_clean %>% 
  ungroup() %>% 
  select(`2014`,`2015`,`2016`, `2017`, `2018`) %>% 
  as.matrix()

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- pums_bart_clean$PUMA
colnames(dist_matrix) <- pums_bart_clean$PUMA

match <- dist_matrix["01309",] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    PUMA = rowname,
    match = "."
  ) %>% 
  right_join(
    pums_bart_clean
  ) %>% 
  arrange(match) %>% 
  .[1:11, ] %>% 
  left_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
  st_as_sf()

```


```{r maps, message=FALSE, warning=FALSE, echo=FALSE}

leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match[1, ],
    color = "red",
    label = ~PUMA
  ) %>% 
  addPolygons(
    data = match[-1, ],
    label = ~PUMA
  )
```

```{r plot, message=FALSE, warning=FALSE, echo=FALSE}

match_pumas <-
  match %>% 
  filter(!PUMA %in% c("01309")) %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>%
  group_by(
    year
  ) %>% 
  summarize(
    bart = mean(bart),
    PUMA = "Similar PUMAs"
  )

treatment_pumas <-
  match %>% 
  filter(PUMA %in% c("01309")) %>% 
  select(-match) %>% 
  st_drop_geometry() %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  )

rbind(
  treatment_pumas,
  match_pumas
) %>% 
  ggplot(
    aes(
      x = as.numeric(year),
      y = bart,
      color = PUMA
    )
  ) +
  geom_line() +
  geom_vline(xintercept = 2018.5, linetype = "dashed") +
  labs(
    title = "Brentwood vs. control neighborhoods, BART ridership",
    x = "Year",
    y = "BART commuters"
  )

```


```{r, message=FALSE, warning=FALSE, echo=FALSE}
transit_did <-
  match %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>% 
  mutate(
    year = year %>% as.numeric(),
    time = ifelse(year > 2018, 1, 0),
    treated = ifelse(PUMA == "01309", 1, 0)
  )

did_reg <- lm(bart ~ treated*time, data = transit_did)

summary(did_reg)
```

# Comments on Model

While there appears to be an increase in BART commuters due to the opening of Antioch Station, this result does not appear to be of statistical significance as the p-value is > 5%. In addition, the high residual standard error of 1304 and the presence of a negative R-square value of -0.0302 suggests that the chosen model under current constraints may not represent the best fit for the available data.

Key notes on the analysis parameters:

1. Replicate weights not adopted to correct for standard error. Incorporating replicate weights in this study could have reduced the standard error for a more representative study.

2. Model above evaluates bart ridership departing from Brentwood, which may not be the most accurate measure to evaluate the effect of Antioch Station. In addition, the PUMAs data registers commuters who use BART for work trips and does not account for non-work trips.

3. Model above evaluates data over a 5 year period between 2014-2019. Ridership on transit projects are typically characterized by minimal increases in the initial years and a more exponential increase as other developments (i.e. housing/commercial) mature within the vicinity of the transit development. As Antioch Station turned operational in mid-2018, an analysis of ridership levels at Brentwood over a longer duration post construction would be more representative.

4. Apart from train ridership, no other control variables were used to identify matching PUMAs. This may have a limiting effect over the study.
 