---
title: "AlexanderNg_A2"
author: "Alexander Ng"
date: " February 1, 2022"
output: html_document
---
 
```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(tigris)
library(censusapi)
library(sf)
library(leaflet)
library(htmltools)
library(tidycensus)
library(mapview)
library(dplyr)
library(ggrepel)
library(ggplot2)

knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning = F, message = F)

```

```{r, include = FALSE}
path <- "/Volumes/GoogleDrive/Shared drives/SFBI/Data Library/NHTS/nhts17-caltrans-tsdc-download/"

pois <- st_read("/Volumes/GoogleDrive/Shared drives/SFBI/Data Library/OSM/gis_osm_pois_a_free_1.shp")

```

# Assignment 2 - Complete Communities
 
## Introduction

This report focuses on applying the "complete communities" methodology within a specific geographic location. The methods of analysis includes deciding on a selected range of amenities that define a "complete community and assigning weights/rankings to these amenities based on the time taken to travel to these amenities using different modes of transportation (i.e. walking, biking, driving).

```{r, include = FALSE}
# pois_summary <- pois %>% 
#   st_drop_geometry() %>% 
#   group_by(fclass) %>% 
#   count() %>% 
#   arrange(desc(n))
# 
# bay_county_names <-
#   c(
#     "Alameda",
#     "Contra Costa",
#     "Marin",
#     "Napa",
#     "San Francisco",
#     "San Mateo",
#     "Santa Clara",
#     "Solano",
#     "Sonoma"
#   )
# 
# bay_counties <-
#   counties("CA", cb = T, progress_bar = F) %>%
#   filter(NAME %in% bay_county_names) %>%
#   st_transform(st_crs(pois))
# 
# bay_pois <- pois %>% 
#   .[bay_counties, ] %>% 
#   rename(amenity = fclass)
# 
# mapview(bay_pois, zcol = "amenity")
# 
# #Let's not use all POIs lets filter down to differentiate
# 
# bay_pois_filter <- bay_pois %>% 
#   filter(amenity %in% c(
#     "park",
#     "supermarket",
#     "pharmacy",
#     "bank",
#     "school",
#     "doctors",
#     "restaurant",
#     "bar",
#     "pub",
#     "pitch",
#     "playground",
#     "hospital",
#     "theatre"
#   ))
# saveRDS(bay_pois_filter, "bay_pois_filter.RDS")
```

```{r, include = FALSE}
# bay_pois_filter2 <- bay_pois %>% 
#   filter(amenity %in% c(
#     "hospital"))
# 
# saveRDS(bay_pois_filter2, "bay_pois_filter2.RDS")
```

```{r, include = FALSE}
# smc_cbs <- blocks("CA","San Mateo")
# 
# rdw_boundary <- places("CA") %>% 
#   filter(NAME == "Redwood City")
# 
# rdw_cbs <- smc_cbs %>% 
#   st_centroid() %>% 
#   .[rdw_boundary, ] %>% 
#   st_drop_geometry() %>% 
#   left_join(smc_cbs %>% select(GEOID10)) %>% 
#   st_as_sf()
# 
# saveRDS(rdw_cbs, "rdw_cbs.RDS")


# Isochrones

# isochrone <- mb_isochrone(
#    rdw_cbs,
#    profile = "walking",
#    time = c(5,10,15)
#  ) %>% mutate(
#  mode = "driving")
#  
#  isochrone[1:3,] %>% mapview()
#  
# saveRDS(isochrone,"isochrone_rwc_walking.RDS")
# 
#  isochrone_2 <- mb_isochrone(
#    rdw_cbs,
#    profile = "driving",
#    time = c(5,10,15)
#  ) %>% mutate(
#  mode = "driving")
#  
#  isochrone_2[1:3,] %>% mapview()
#  
#  saveRDS(isochrone_2, "isochrone_rwc_driving.RDS") 
# 
#  isochrone_3 <- mb_isochrone(
#    rdw_cbs,
#    profile = "cycling",
#    time = c(5,10,15)
#  ) %>% mutate(
#    mode = "biking")
# 
#  isochrone_3[1:3,] %>% mapview()
#  
#  saveRDS(isochrone_3, "isochrone_rwc_biking.RDS")

```

Firstly, Redwood City in San Mateo County was selected as the region of study due to its reputation for a high standard of living here in the Bay Area. The isochrones used for this analysis were determined using census block data (and not census block groups as demonstrated in the class example).

In determining the Points-of-Interests (POIs) from OpenStreetMap for this study, reference was taken from quality of life surveys carried out in the US and POI selection informed through amenities that respondents indicated they would like to see more (Quality of Life, 2019) such as parks, recreational facilities, trails, restaurants and theaters. Further reference was taken from the Walk Score indicator which listed amenities such as grocery stores and retail forming a higher walk score (Walk Score, 2022). For completeness,  amenities essential for maintaining a baseline quality of life such as supermarkets and access to medical care. The resulting POIs were therefore selected to develop our complete community baseline:

1. Park
2. Supermarket
3. Pharmacy
4. Bank
5. School
6. Doctors Offices
7. Restaurant
8. Bar
9. Pub
10. Pitch
11. Playground
12. Hospital
13. Theatre


```{r, include = FALSE}
bay_pois_filter <- readRDS("bay_pois_filter.RDS")
bay_pois_filter2 <- readRDS("bay_pois_filter2.RDS")

# mapview(bay_pois_filter2, zcol = "amenity")
```

## Map of Selected POIs

A map of the selected POIs is illustrated in the following figure:
```{r}

mapview(bay_pois_filter, zcol = "amenity")
```

## Map of Redwood City Census Blocks

A map of Redwood City Census Blocks is illustrated in the following figure:
```{r, include = FALSE}
rdw_cbs <- readRDS("rdw_cbs.RDS")

```

```{r}
mapview(rdw_cbs)
```



```{r, include = FALSE}
devtools::install_github("walkerke/mapboxapi")

library(mapboxapi)

mb_access_token("sk.eyJ1IjoiYWxleG5nbGQiLCJhIjoiY2t5dnY3OGliMDJlaTMwcnIxeHd5bTJucyJ9.mZzy6K59_f5dGUlEQ8oL7Q", install = T, overwrite = T)
```

## Isochrones

We established separate isochrones for the walking, biking and driving in 5,10 and 15 min time buffers for the purpose of this analysis. While this could also be carried out using the map-dfr function to streamline the analysis, the isochrones were kept separate to facilitate individual team members to carry out the analyses separately before consolidating the results using saveRDS and rbind. The combined isochrones are mapped below:

```{r, include = FALSE}

isochrone <- readRDS("isochrone_rwc_walking.RDS")
isochrone_2 <- readRDS("isochrone_rwc_driving.RDS")
isochrone_3 <- readRDS("isochrone_rwc_biking.RDS")

isochrone_combined <- rbind(isochrone, isochrone_2, isochrone_3)

```

```{r}
mapview(isochrone_combined)
```


## Complete Community Baseline

To develop the community scores, subjective weights using a decay function were created based on relative distance and minimum quantity of each amenity type. A similar process was used to assign weights for the three distinct modal types: Walking, Biking and Driving. 

```{r, include = FALSE}

sf::sf_use_s2(FALSE)
access_raw <- isochrone_combined %>% 
  st_make_valid() %>%
  st_join(bay_pois_filter) %>% 
  st_drop_geometry()

```

```{r, include=FALSE}
amenity_preference <- data.frame(
  amenity = c(
    "park",
    "supermarket",
    "pharmacy",
    "bank",
    "school",
    "doctors",
    "restaurant",
    "bar",
    "pub",
    "pitch",
    "playground",
    "hospital",
    "theatre"
  ),
  amenity_value = c(
    0.9,
    1,
    0.5,
    0.5,
    0.8,
    0.6,
    0.7,
    0.7,
    0.7,
    0.8,
    0.8,
    0.9,
    0.3
  ),
  amenity_quantity = c(
   2,
   2,
   1,
   1,
   1,
   2,
   20,
   5,
   5,
   10,
   5,
   1,
   2
  )
) %>% 
  mutate(
    amenity_decay = -log(0.5)/(amenity_quantity)
  )

mode_preference <- data.frame(
  mode = c(
    "walking",
    "biking",
    "driving"
  ),
  mode_value = c(
    1,
    0.5,
    0.7
  ),
  mode_reasonable = c(
    15,
    15,
    25
  )
) %>% 
  mutate(
    mode_decay = -log(0.5)/mode_reasonable
  )
```

```{r}
amenity_preference
mode_preference
```


```{r, include=FALSE}
complete_baseline <- data.frame(
  amenity = amenity_preference$amenity %>% 
    rep(amenity_preference$amenity_quantity)
) %>% 
  left_join(
    amenity_preference,
    by = "amenity"
  ) %>% 
  group_by(amenity) %>% 
  mutate(
    amenity_rank = row_number() - 1
  ) %>% 
  ungroup() %>% 
  mutate(
    score = amenity_value * exp(-amenity_rank * amenity_decay) * 0.5
  ) # ASSUMES ALL 0.5, or mode_factor for 15 min of any transport means


```

```{r}
sum(complete_baseline$score)
```

To form a reference point for the different scores, a community baseline was used that assumes that all amenities were 15 minutes away and each amenity had the same number of instances as the amenity quantity. The baseline value were then used to normalize the scores by dividing the calculated values against the baseline. Using the techniques described in class, we were able to derive our complete baseline score of 16.11. 

The completeness score was then determined by multiplying the amenity value, amenity rank, and mode value, and applying a decay function to both the amenity rank and mode value.   

```{r, include = FALSE}
complete_temp <- access_raw %>%
  left_join(
    amenity_preference,
    by = "amenity"
  ) %>%
  select(-amenity_quantity) %>%
  left_join(
    mode_preference,
    by = "mode"
  ) %>%
  select(-mode_reasonable) %>% 
  group_by(id,mode,amenity) %>%
  arrange(time) %>%
  mutate(
    amenity_rank = row_number() - 1
  ) %>% 
  ungroup()
```

```{r, include = FALSE}
 complete_modes <- complete_temp %>% #for one mode only
    mutate(
      score = amenity_value *exp(-amenity_rank * amenity_decay) * exp(-time * mode_decay)
    ) %>%
    group_by(id, mode) %>%
    arrange(desc(score)) %>%
    filter(!duplicated(osm_id)) %>%
    summarize(
      score = sum(score, na.rm=T)/sum(complete_baseline$score)
    )
```

```{r,include=FALSE}
complete_total <- complete_temp %>%
  mutate(
      score = amenity_value *exp(-amenity_rank * amenity_decay) * mode_value * exp(-time * mode_decay)
    ) %>%
  group_by(id) %>%
  arrange(desc(score)) %>%
  filter(!duplicated(osm_id)) %>%
  summarize(
    score = sum(score, na.rm = T)/sum(complete_baseline$score)
  ) %>%
  mutate(mode = "total")

complete <- rbind(
  complete_modes,
  complete_total
)

complete_map <- complete %>%
  pivot_wider(
    names_from = "mode",
    values_from = "score"
  ) %>%
  cbind(rdw_cbs %>% select(GEOID10)) %>%
  st_as_sf()

```

## Map of Complete Community Score in Redwood City
```{r}
mapview(complete_map, zcol = "total")
```

The complete community analysis for Redwood City indicates that amenity distribution in Redwood City is relatively uniform. In general, blocks located closest to main arterial corridors such as Bayshore Freeway have the highest community scores due to better connectivity and access and the scores decrease for block groups located farther away from the downtown core. 

# EQUITY ANALYSIS

The study takes a closer examination at the racial distribution relative to the median complete community score for Redwood City using Decennial census block data. The results are represented in the following charts that show the distribution of racial demographic groups that were either above or below the median community score. 

```{r}

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

dec_vars_2020 <-
  listCensusMetadata(
    name = "2020/dec/pl",
    type = "variables"
  )

rwc_pop_race_2020 <-
  getCensus(
    name = "dec/pl",
    vintage = 2020,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "group(P1)"
  ) %>% 
  mutate(
    block =
      paste0(state,county,tract,block)
  ) %>% 
  select(!c(GEO_ID,state,county,tract,NAME) & !ends_with(c("NA"))) %>% 
  pivot_longer(
    ends_with("N"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    dec_vars_2020 %>% 
      select(name, label)
  ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,NA,"category1","category2"),
    sep = "!!"
  ) %>%
  left_join(
    rdw_cbs %>%
      select("GEOID10"),
    by = c("block"="GEOID10") 
  ) %>%
  select(-category1)

score_median = median(complete_map$total) 
  
rwc_filter <- rwc_pop_race_2020 %>%
  filter(!is.na(category2)) %>%
  left_join(
    complete_map %>%
      select("total"),
    by = 'geometry'
  ) %>%
  filter(!is.na(total)) %>%
  mutate(
    Threshold = ifelse(total>score_median,"Above Median","Below Median")
  ) %>%
  mutate(
    race = case_when(
      category2 %in% c("White alone")~"White",
      category2 %in% c("Black or African American alone")~"Black or African American",
      category2 %in% c("American Indian and Alaska Native alone")~"American Indian and Alaska Native",
      category2 %in% c("Asian alone")~"Asian",
      category2 %in% c("Native Hawaiian and Other Pacific Islander alone")~"Native Hawaiian and Other Pacific Islander",
      category2 %in% c("Some Other Race alone", "Population of two races:", "Population of three races:" , "Population of four races:" , "Population of five races:",  "Population of six races:")~"Other"
    )
  ) %>%
  group_by(
    Threshold, race
  ) %>%
  summarize(
   estimate=sum(estimate)
  ) %>%
  group_by(
    Threshold
  ) %>%
  mutate(
    perc = estimate/sum(estimate) *100
  )

df2 <- rwc_filter %>% 
  mutate(csum = rev(cumsum(rev(perc))), 
         pos = perc/2 + lead(csum, 1),
         pos = if_else(is.na(pos), perc/2, pos))

rwc_filter %>% 
  ggplot(data = rwc_filter,
         mapping = aes(x = "" , 
             y = perc, 
             fill = fct_inorder(race)
         )) +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~Threshold) +
  scale_fill_brewer(palette = "Pastel1") +
  geom_label_repel(data = df2,
                  mapping = aes(y = pos, label = paste0(round(perc), "%")),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Completeness Score by Race")) +
  theme_void()

```

Overall, there is a higher proportion of White/Asian only demographic groups that are below the median community score. While this may be due to the higher proportion of White and Asian population in Redwood City in general, including additional information such as household income could provide more insights for the current distribution trend.

# CRITICAL AMENITY ANALYSIS

Finally, a critical amenity analysis was carried out to determine the impact of adding a new critical amenity such as a hospital. This was determined by filtering for hospitals only and using driving isochrone data to identify blocks that were either within or outside a 10 minute access to a hospital (Yes or No). The results of this analysis are as demarcated:

```{r, include = FALSE}
access_raw2 <- isochrone_2 %>% 
  st_make_valid() %>%
  st_join(bay_pois_filter2) %>% 
  st_drop_geometry() %>%
  filter(!is.na(osm_id)) %>%
  group_by(
    id
  ) %>%
  mutate(
    mintime = min(time)
  ) %>%
  left_join(
    complete_map%>%
      select(id,geometry),
    by="id"
  ) %>%
  filter(!duplicated(id)) %>%
  mutate(
    Access = ifelse(mintime>10,"No","Yes")
  ) %>%
  st_as_sf()

```

```{r}
mapview(access_raw2, zcol="Access")
```

## Constructing a New Hospital

Given the current hospital access distribution, further analysis was carried out to determine the impact of constructing a new hospital along Bayshore Freeway in the community. This was determined by manually adding a new hospital POI and observing the changes in critical amenity access distribution. The location of the new hospital is indicated in the following map:

```{r, include = FALSE}
new_hospital <- data.frame(
  osm_id=00000001, code = 2110, amenity = "hospital", name = "New Hospital")

newhospital_point <- as.data.frame(37.48956) %>%
  rename(lat = "37.48956") %>%
  mutate(lon = -122.2239) %>%
  sf::st_as_sf(., coords = c("lon", "lat"), crs = 4326)

new_hospital <- new_hospital %>%
  cbind(newhospital_point) %>%
  st_as_sf(crs=4326)

bay_pois_filternew <- bay_pois_filter2 %>%
  rbind(new_hospital)

mapview(bay_pois_filternew)

```

```{r}
mapview(new_hospital)
```


```{r}

access_raw3 <- isochrone_2 %>% 
  st_make_valid() %>%
  st_join(bay_pois_filternew) %>% 
  st_drop_geometry() %>%
  filter(!is.na(osm_id)) %>%
  group_by(
    id
  ) %>%
  mutate(
    mintime = min(time)
  ) %>%
  left_join(
    complete_map%>%
      select(id,geometry),
    by="id"
  ) %>%
  filter(!duplicated(id)) %>%
  mutate(
    Access = ifelse(mintime>10,"No","Yes")
  ) %>%
  st_as_sf() 

```

```{r}
mapview(access_raw3, zcol="Access")

```

The analysis indicates that constructing a new hospital at our proposed location allows a significantly larger proportion of Redwood City residents to access the new facility as compared to the existing distribution. This analysis could be useful in determining whether proposed amenities in the future could help address underserved regions with low community scores.

# CONCLUSION

The "Complete Community" approach used in this study is a useful benchmark for assessing amenity distribution in different geographic locations and provides ample flexibility for different analyses to be carried out such as equity analysis and impact assessment. Nevertheless, there are limitations as the methodology used are reliant on the use of weightings and rankings that are subjective and may not be representative of actual ground sentiment. Therefore, the "Complete Community" methodology should be complemented with other community studies such as surveys and interviews for more holistic community building. In addition, there are other tools available such as "WalkScore" API which allows users to assess a given location's WalkScore or TransitScore based on its relative distance to amenities/transport facilities.

# REFERENCES USED

1. Quality of Life Survey: http://www.kcqualityoflife.org/
2. WalkScore API: https://www.walkscore.com/professional/api.php
  