---
title: "AlexanderNg_A3"
author: "Alexander Ng"
date: "10/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r temp files, message=FALSE, warning=FALSE, echo=FALSE}
temp <- tempfile()
download.file("https://www2.census.gov/programs-surveys/acs/data/pums/2019/1-Year/csv_hca.zip",destfile = temp)

pums_hca_2019_1yr <- read_csv(unzip(temp,"psam_h06.csv"))

unlink(temp)
```

```{r pull up data, message=FALSE, warning=FALSE, echo=FALSE}

pums_2019_1yr <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2019,
  region = "public use microdata area:*", 
  regionin = "state:06",
  vars = c(
    "SERIALNO",
    "SPORDER",
    "PWGTP",
    "WGTP",
    "NP",
    "HHL",
    "HINCP",
    "TEN",
    "AGEP",
    "DDRS",
    "HICOV"
  )
)

```

```{r filter relevant fields, message=FALSE, warning=FALSE, echo=FALSE}

ca_pumas <-pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

bay_pums <-
  pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0")
  ) %>% 
  filter(PUMA %in% bay_pumas$PUMACE10)

```

```{r identify household specific criteria based on susceptibility to heat injuries, message=FALSE, warning=FALSE, echo=FALSE}

bay_pums_heat <-
  bay_pums %>% 
  mutate(
    PWGTP = as.numeric(PWGTP),
    elder_nohealthinsurance = ifelse(
      (AGEP >= 65) & #select elderly aged 65 & over
        (HICOV == 2), #select no health insurance coverage
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA) %>% 
  summarize(
    sum_elder_nohealthinsurance = sum(elder_nohealthinsurance, na.rm = T),
    sum_perc_elder_nohealthinsurance =
      sum(elder_nohealthinsurance, na.rm =T)/sum(PWGTP, na.rm = T)*100
  ) %>% 
  left_join(
    bay_pumas %>% 
      select(PUMACE10),
    by = c("PUMA" = "PUMACE10")
  ) %>% 
  st_as_sf()

```

```{r sum, message=FALSE, warning=FALSE, echo=FALSE}
sum <- sum(bay_pums_heat$sum_elder_nohealthinsurance)
```


```{r plot map, message=FALSE, warning=FALSE, echo=FALSE}
pums_pal <- colorNumeric(
  palette = "Oranges",
  domain = bay_pums_heat$sum_elder_nohealthinsurance
)

leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data = bay_pums_heat,
    fillColor = ~pums_pal(sum_elder_nohealthinsurance),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(sum_elder_nohealthinsurance), 
      " No. of Elderly & no health insurance"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = bay_pums_heat,
    pal = pums_pal,
    values = ~sum_elder_nohealthinsurance,
    title = " No. of Elderly & no health insurance"
  )

```

##General Observations

The above data map tracks elderly population in the Bay Area who do not have any health insurance records as this is the community identified to be the most vulnerable to heat dangers. There are approximately 16,833 observed individuals who classify as such in 2019.

From the heat map, it appears that there is a wide geographical spread of these vulnerable individuals and the majority of whom are residing in the Bay Area suburbs (exception of a small cluster of vulnerable elderly residing within SG city). The wide geographical range across suburbs that are not well-served by highway infrastructure may pose some logistical planning challenges, particularly for emergency response teams who are responding to heat-related emergencies in the Bay Area. The current spread also suggests that city council officials may wish to consider setting up more temporary housing shelters in the suburban regions of the Bay Area so that vulnerable individuals can be evacuated more readily in the event of heat emergencies.
